---
title: "Seasonal ARIMA models"
author: "Kevin W. McConeghy"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{flumodelr: 06-arima}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(tibble.print_min = 4L, tibble.print_max = 4L)
library(flumodelr)
library(lubridate)
library(scales)
library(forecast)
```

# Autoregressive models  

When modeling time series in linear predictor model, past values of dependent 
variable can be included as linear predictors in what is termed an 
autoregression model (AR). The number of lagged dependent variables used as 
predictors is referred to as the 'order' of an autoregressive model.

#### Example. AR(2) Model  

$$AR:  y_t = \alpha_0 + \beta_1*y_{t-1} + \beta_2*y_{t-2} + \varepsilon$$

# Moving Average Models  

An alternative approach is to use a moving average model. 

$$MA: y_{t} = \alpha_0 + \beta_1*\varepsilon_{t-1} + \beta_2*\varepsilon_{t-2}
+ \dots + \beta_q*\varepsilon_{t-q}$$
An MA(q) model uses past forecasting (fit) errors in a regression model.  

Both models can be combined into what is called an 'ARIMA' model or autoregressive 
integrated moving average.  

$$ARIMA: \ y'_{t} = \alpha_0 + \beta_1*y'_{t-1} + \beta_2*y'_{t-2} + \dots +
\beta_p*y'_{t-p} + \theta_1*\varepsilon_{t-1} + \theta_2*\varepsilon_{t-2}
+ \dots + \theta_q*\varepsilon_{t-q} + \varepsilon{_t}$$

The ' signifies differencing (i.e. difference between forecasted points). A model
is referred to as ARIMA (p, d, q), where p = AR order, d = degree of differencing, 
q = order of MA.  

Models can be specified or fit automatically using AIC or other fit values.  

To determine p and q values researchers often plot autocorrelation.  

### Exampling influenza via ACF / PACF plots  
```{r }
fludta <- flumodelr::fludta

flu_deaths <- fludta$fludeaths
```

An 'ACF' or autocorrelation function plot computes the correlation between y_t and
y_t-1, ..., ty-p. 
```{r }
ggAcf(flu_deaths, main='Autocorrelation plot')
```
Note because influenza highly peaked by season, and the data contains more than 
one season you see the correlation flip over time.  

However, a limitation with ACF plots is that if timepoint y_t and y_t-1 are correlated 
then y_t and y_t-2 may be also simply because they are near y_t-1. 

To account for this individuals use partial autocorrelation plots.
```{r }
ggPacf(flu_deaths, main='Partial autocorrelation plot')
```
The decaying in the ACF along with the two initial spikes in the PACF suggest
a p order of at least 2 is necessary, but maybe not a MA model. 

```{r }
arima(flu_deaths, order=c(2, 0, 0))

```
Compare this to auto.arima, which performs an automatic search for the correct
order. 

```{r }
auto.arima(flu_deaths, stepwise=F, approximation=F)
```
The auto.arima model suggests a MA(4) is appropriate, with slightly lower AIC.  

A useful summary function is `ggtsdisplay`.  
```{r }
ggtsdisplay(flu_deaths)
```

A full review of ARIMA modelling should be sought elsewhere.  

# Seasonal ARIMA  

Influenza disease is not stationary over time, it is highly seasonal spiking 
in the winter months. 
```{r }
flu_deaths %>% diff() %>% ggtsdisplay(., main='First Difference, yt\'')
```

A specific set of ARIMA models can be used in this case.
```{r }
fit <- Arima(flu_deaths, order=c(1, 0, 4), seasonal=c(0,1,1))
fit
```
You can also plot the residuals:
```{r }
fit %>% residuals() %>% ggtsdisplay()
```
Notice the significant spike at lag 5.

Now with Arima(0, 0, 5)(0,1,1):
```{r }
Arima(flu_deaths, order=c(0, 1, 5), seasonal=c(0,1,1)) %>% residuals() %>% ggtsdisplay()
```
Better.  

```{r }
Arima(flu_deaths, order=c(0, 1, 5), seasonal=c(0,1,1)) %>% checkresiduals()

```

In flumodelr, `flum` or `flurima` will take given ARIMA models or use auto.arima 
to fit if none are specified.  

```{r }
fit <- flurima(fludta, outc = fludeaths, time=week_in_order, echo=T)
```

# References  

```{r }
sessioninfo::session_info()
```
