---
title: "Imputation for Baseline data"
author: "Kevin W. McConeghy"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{flumodelr: 10-impute}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(tibble.print_min = 4L, tibble.print_max = 4L)
library(flumodelr)
library(lubridate)
library(scales)
library(forecast)
library(imputeTS)
```

# Imputation

Constructing the counterfactual of baseline influenza activity during an active influenza season can be put into the framework 
of missing data. 

The influenza season time of yyyy-40 -> yyyy-23 can be set to missing, then using imputation methods you can fill in the missing timepoints using the off-season observations.  

The `ImputeTS` package allows for a variety imputation approaches.  

### Influenza data  
```{r }
fludta <- flumodelr::fludta
fludta
```

## Create *missing* dataset  
```{r }
fludta_miss <- if_else(fludta$epi, NA_integer_, fludta$fludeaths)
```

## kalman  
```{r }
fludta_im <- na.kalman(fludta_miss, model='auto.arima')
```

```{r }
mydate <- as_date(fludta$yrweek_dt)
fludta_plot <- cbind(date = mydate, 
                         fludta_miss,
                         fludta_im) %>% as.data.frame(.)

ggplot(data=fludta_plot, aes(x=date)) +
  geom_point(aes(y=fludta_miss)) +
  geom_line(aes(y=fludta_im))
```

## `flugam` function  

### Using only off-season as baseline  

```{r }
fit <- fludta %>%
  bind_cols(., flugam(data=fludta, outc=fludeaths, time=week_in_order, season=epi))
```

```{r }
#Set up graph labels, line specs
line_names <- c("Deaths From P&I", "Expected", "Baseline")
line_cols <- c("#CC0000", "black", "black")
line_types <- c(1, 1, 2)
names(line_cols) <- line_names
names(line_types) <- line_names

ggplot(fit, aes(x=yrweek_dt)) + 
  geom_line(aes(y=fludeaths, colour=line_names[[1]], 
                linetype=line_names[[1]]), size=0.8) +
  geom_line(aes(y=flu_pred, colour=line_names[[2]], 
                linetype=line_names[[2]]), size=0.8) +
  geom_line(aes(y=flu_base_upr, colour=line_names[[3]], 
                linetype=line_names[[3]]), size=0.8) +
  scale_colour_manual("Line", breaks=line_names, values = line_cols) +
  scale_linetype_manual("Line",  breaks=line_names, values = line_types) +
  scale_x_date(labels = date_format("%Y"), date_breaks="1 year",
               expand=c(0, .9)) + 
  xlab("Year") + 
  ylab("% of Deaths from P&I") + 
  theme_light(base_size=14) +
  theme(legend.text=element_text(size=10), 
        plot.title = element_text(size=14)) +
  labs(title="Figure 1. Pneumonia and Influenza Mortality",
       caption="Generalized additive model - Seasonal only") +
  guides(colour = guide_legend("Line"), linetype = guide_legend("Line"))

```

### Using viral data as baseline  
```{r }
fit <- fludta %>%
  bind_cols(., flugam(data=fludta, outc=fludeaths, 
                      time=week_in_order, viral='prop_flupos'))
```

```{r }
#Set up graph labels, line specs
line_names <- c("Deaths From P&I", "Expected", "Baseline")
line_cols <- c("#CC0000", "black", "black")
line_types <- c(1, 1, 2)
names(line_cols) <- line_names
names(line_types) <- line_names

ggplot(fit, aes(x=yrweek_dt)) + 
  geom_line(aes(y=fludeaths, colour=line_names[[1]], 
                linetype=line_names[[1]]), size=0.8) +
  geom_line(aes(y=flu_pred, colour=line_names[[2]], 
                linetype=line_names[[2]]), size=0.8) +
  geom_line(aes(y=flu_base_upr, colour=line_names[[3]], 
                linetype=line_names[[3]]), size=0.8) +
  scale_colour_manual("Line", breaks=line_names, values = line_cols) +
  scale_linetype_manual("Line",  breaks=line_names, values = line_types) +
  scale_x_date(labels = date_format("%Y"), date_breaks="1 year",
               expand=c(0, .9)) + 
  xlab("Year") + 
  ylab("% of Deaths from P&I") + 
  theme_light(base_size=14) +
  theme(legend.text=element_text(size=10), 
        plot.title = element_text(size=14)) +
  labs(title="Figure 2. Pneumonia and Influenza Mortality",
       caption="Generalized additive model - Viral added") +
  guides(colour = guide_legend("Line"), linetype = guide_legend("Line"))

```

# References  

```{r }
sessioninfo::session_info()
```
