---
title: "Create a simulated influenza season"
author: "Kevin W. McConeghy"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{flumodelr: 11-simdata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(tibble.print_min = 4L, tibble.print_max = 4L)
library(flumodelr)
library(lubridate)
library(scales)
library(prophet)
```

Basic workflow for creating a 'simulation' influenza season. 

1) Use CDC reported data to inform general shape / curve.

  1a) Integrate the CDC P&I deaths into a density plot.  
  1b) Fit a two-mixture beta regression to determine initial values.  

2) Use beta regression parameters as a guide for simulation  
  2a) Simulate season density using two-mixture beta distribution.
  2b) Convert density back to raw counts using arbitrary value.
  
3) Refine as a function program  

# STEP 1: empirical influenza season  

## CDC data (see ??)  
```{r, echo=FALSE}
EpiWeekToDates <- function(year, weeknum) {
  
  #Compute 4th day in January
    jan4 <- ymd(paste(year, 1, 4, sep="-"))
  
  #Compute day of week
    DofW <- wday(jan4, week_start = 7)-1 #Sunday start
    #The -1 is  correction for how wday() counts days  
    
  #If Jan 4th was Sun, starting week date is 01/04/xxxx
    startweek <- if_else(DofW == 7, jan4, (jan4 - (DofW)))
 
  #First Date of Week
    d0 = startweek + (weeknum - 1) * 7
  
  #Last Date of Week
    d1 = startweek + (weeknum - 1) * 7 + 6 
  
  return(list(d0 = ymd(d0), d1 = ymd(d1)))
}
```

```{r, echo=F, results='asis', fig.width=7}
cdc_122 <- flumodelr::cdc122city

#some formatting 
cdc_122 <- cdc_122 %>%
  group_by(year, week) %>%
  summarize(fludeaths = sum(deaths_pnaflu, na.rm=T)) %>%
  mutate(week_dt = EpiWeekToDates(year, week)[[1]])
```

```{r }
g1 <- ggplot(cdc_122, aes(x=week_dt, y=fludeaths)) + 
  geom_line(size=0.8, colour="#CC0000") +
  scale_x_date(labels = date_format("%Y"), date_breaks="1 year",
               expand=c(0, .9)) + 
  xlab("Year") + 
  ylab("No. of Deaths from P&I") + 
  theme_light(base_size=16) +
  theme(plot.title = element_text(size=14)) +
  labs(title="Figure 1. Pneumonia and Influenza Mortality") +
  theme(axis.text.x=element_text(angle=90, hjust=1))
g1
```
Because there is a systematic shift in the 1990s, will start modeling after that point.  

```{r }
cdc_122_1990 <- dplyr::filter(cdc_122, year>=1990)
```

```{r }
g1 <- ggplot(cdc_122_1990, aes(x=week_dt, y=fludeaths)) + 
  geom_line(size=0.8, colour="#CC0000") +
  scale_x_date(labels = date_format("%Y"), date_breaks="1 year",
               expand=c(0, .9)) + 
  xlab("Year") + 
  ylab("No. of Deaths from P&I") + 
  theme_light(base_size=16) +
  theme(plot.title = element_text(size=14)) +
  labs(title="Figure 1. Pneumonia and Influenza Mortality") +
  theme(axis.text.x=element_text(angle=90, hjust=1))
g1
```
Much more stable.  

## Prophet fit
```{r }
fludta_phet <- cdc_122_1990 %>%
  dplyr::select(week_dt, fludeaths) %>%
  dplyr::rename(ds = week_dt, y=fludeaths)

p_fit <- prophet(fludta_phet, 
                 seasonality.mode = 'multiplicative',
                 mcmc.samples = 2000,
                 interval.width = 0.95)
```

```{r }
future <- make_future_dataframe(p_fit, periods = 53, freq='week')
tail(future)
```
```{r }
forecast <- predict(p_fit, future)
tail(forecast[c('ds', 'yhat', 'yhat_lower', 'yhat_upper')])
```

```{r }
plot(p_fit, forecast)
```
Prophet does not fit the outlier seasons well (as expected they were
unusually severe season).

## Data generating process  

### Total deaths  
Total deaths obtained from actual CDC reported data for 2010-2015. It is assumed
this value is accurate, but what is unknown are the proportion of deaths 
due to influenza.  
```{r }
flu_real <- fludta %>%
  dplyr::filter(week<=52)
```

```{r }
ggplot(flu_real, aes(x=week_in_order, y=alldeaths)) +
  geom_line()
```

We add the real all-cause deaths to the dataset.  
```{r }
flu_sim$deaths_all <- flu_real$alldeaths
```

The objective is to simulate deaths attributable to influenza under different 
assumptions.  

A seasonal influenza data generating process.  
Features of influenza deaths:
1) Mild or severe season, qualitatively increases number of deaths. 
2) Peak month, which month has peak   
3) Peakedness, whether a sharp rise/fall, or gentle slope.  

## Peakedness simulated with vertex parabola  
$$Eq \ 1. \  y = a(x-h)^2 + k$$  
Where `k` is the highest observed influenza death, `h` is the peak week, and `a` 
determines peakedness.  

```{r }
#y = a(x-h)^2 + k

```

```{r }
#peak of february, week 4
flu_sim_1 <- flu_sim %>%
  mutate(peak_week = row_number() - 4, 
         peak = dgamma(week, shape = 4.2, scale=1.2),
         deaths_flu_t = round(deaths_all*peak * season))


```

