---
title: "Incidence rate-difference models"
author: "Rob van Aalst"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{flumodelr: 05-Incidence rate-difference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(tibble.print_min = 4L, tibble.print_max = 4L)
library(dplyr)
library(ggplot2)
library(lubridate)
library(scales)
library(flumodelr)
library(tibble)
```

```{r, results='asis', fig.width=7}
flu_ex <- flumodelr::flu_ex
```

## Virology data  

In current methods, most researchers incorporate data on the percent of isolates positive with influenza viral types. This is publically available in many cases see `?nrevss`. The data describes the proportion of isolates tested, which may be obtained from participating hospital and outpatient settings. Using this surrogate time-periods of high flu activity can be used to inform spikes in influenza activity, increasing confidence that deaths are influenza related and estimated influenza morbidity and mortality.  

## Definition of the Peri- and Summer-season

Incidence rate-difference models have been used frequently to estimate influenza-associated hospitalizations and
deaths. We followed Thompson and defined five periods for each respiratory season: (i) a period when ‡10% of specimens tested were positive for influenza, (ii) a period when ‡15% of specimens were positive, (iii) a peri-season baseline period when <10% of specimens were positive, (iv) a peri-season baseline period when <15% of specimens were positive, and (v) a summer-season baseline period. The summer-season baseline period was defined as the weeks from July through September at the beginning of each season and May and June at the end of each season when there is little influenza activity.

```{r, echo=F, results='as.is', fig.width=7.0}

```


See [@Thompson2009]