---
title: "Test different models"
author: "Kevin W. McConeghy"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{flumodelr: 10-impute}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(tibble.print_min = 4L, tibble.print_max = 4L)
library(flumodelr)
library(lubridate)
library(scales)
```

# Create simulated dataset  
First step, create a dataset, five years of influenza data in CDC format/canon.   
```{r }
#some have 53 weeks, but ignored for simulated dataset  
year <- rep(seq(2010, 2014, by=1), 52) 
week <- rep(seq(1, 52, by=1), 5) 

flu_sim <- tibble(year, week)
flu_sim <- mutate(flu_sim, 
                  week_in_order = row_number(),
                  season = if_else(week>=40 | week<20, T, F))
```

## Data generating process  

### Total deaths  
Total deaths obtained from actual CDC reported data for 2010-2015. It is assumed
this value is accurate, but what is unknown are the proportion of deaths 
due to influenza.  
```{r }
flu_real <- fludta %>%
  dplyr::filter(week<=52)
```

```{r }
ggplot(flu_real, aes(x=week_in_order, y=alldeaths)) +
  geom_line()
```

We add the real all-cause deaths to the dataset.  
```{r }
flu_sim$deaths_all <- flu_real$alldeaths
```

The objective is to simulate deaths attributable to influenza under different 
assumptions.  

A seasonal influenza data generating process.  
Features of influenza deaths:
1) Mild or severe season, qualitatively increases number of deaths. 
2) Peak month, which month has peak   
3) Peakedness, whether a sharp rise/fall, or gentle slope.  

## Peakedness simulated with vertex parabola  
$$Eq \ 1. \  y = a(x-h)^2 + k$$  
Where `k` is the highest observed influenza death, `h` is the peak week, and `a` 
determines peakedness.  

```{r }
#y = a(x-h)^2 + k

```

```{r }
#peak of february, week 4
flu_sim_1 <- flu_sim %>%
  mutate(peak_week = row_number() - 4, 
         peak = dgamma(week, shape = 4.2, scale=1.2),
         deaths_flu_t = round(deaths_all*peak * season))


```

